An Introduction to Rocker: Docker Containers for R

### 2017-06-22 manuscript.Rmd
WIP

---
title: "An Introduction to Rocker: Docker Containers for R"
author:
  - name: Carl Boettiger
    affiliation: UC Berkeley 
    address:
    - ESPM Department, University of California,
    - 130 Mulford Hall Berkeley, CA 94720-3114, USA
    email:  cboettig@berkeley.edu
  - name: Dirk Eddelbuettel
    affiliation: Debian and R Projects
    address:
    - line 1
    - line 2
    email:  edd@debian.org
abstract: >
  We describe the Rocker project, which provides a widely used suite of Docker images
  with customized R environments for particular tasks. We discuss how this suite is
  organized and how these tools can increase portability, scaling, reproducibility,
  and convenience of R users and developers.
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---

@@

@@

1. These Dockerfiles require occassional minor maintenance to ensure successful builds
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users.

### debian-stable stack

This stack emphasizes stability and reproducibility of the Docker build.

@@

## The r-base image

The r-base image is the official, verified Docker image for R.  Written and maintained by the Rocker developers on behalf of the Docker project, this image is built and hosted directly by Docker's own developers, who also approve any changes or updates to the image.

example use.

Link to further details in Rocker wiki

## The r-devel images

These images provide both the current and under-development versions of R, including images with particular compiler option configurations (UBSAN, clang).  Package authors submitting R packages to the Central R Archive Network (CRAN) are expected to test their packages on both current and development versions of R prior to submission.

## rocker-versioned stack

example use

screenshot

Custom configurations

Link to further details in Rocker wiki

## Additional images

## Metrics of use

@@

This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](http://journal.r-project.org/latex/RJauthorguide.pdf).

>>>>

---
title: "An Introduction to Rocker: Docker Containers for R"
author:
  - name: Carl Boettiger
    affiliation: UC Berkeley 
    address:
    - ESPM Department, University of California,
    - 130 Mulford Hall Berkeley, CA 94720-3114, USA
    email:  cboettig@berkeley.edu
  - name: Dirk Eddelbuettel
    affiliation: Debian and R Projects
    address:
    - line 1
    - line 2
    email:  edd@debian.org
abstract: >
  We describe the Rocker project, which provides a widely used suite of Docker images
  with customized R environments for particular tasks. We discuss how this suite is
  organized and how these tools can increase portability, scaling, reproducibility,
  and convenience of R users and developers.
preamble: >
   \usepackage{longtable}
   \usepackage{svg}
output: 
  rticles::rjournal_article:
        latex_engine: lualatex
---

@@

- GitHub, Docker Hub (Docker Cloud)

 Automated builds on Docker Hub help ensure the integrity of the build process and ensure that the image build corresponds exactly to Dockerfile on GitHub.

@@

1. These Dockerfiles require occassional minor maintenance to ensure successful builds
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users.
4. `testing` and `unstable` do not recieve the same attention to security updates.

### debian-stable base

This stack emphasizes stability and reproducibility of the Docker build.  The key feature of this stack is the ability to run older versions of R, along with the contemporaneous versions of R packages.  A user specifies the version desired using an image tag, e.g. 
`rocker/r-ver:3.3.1` will refer to an image with R version 3.3.1 installed.  Omitting the tag is equivalent to using the tag `latest`, which, as the name implies, will always point to an image using the current R release.  Thus, users wanting to create downstream Dockerfiles which are based on the current release at time but will continue to reconstruct the same environment in the future after newer R versions are released should explicilty include the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time of writing, and not the `latest` tag.  Users can also run the current development version of R using the tag `devel`, which is built nightly from R-devel sources from `subversion`.

**MRAN archives**: To facilitate installation of only contemporaneous versions of R packages on these images, the default CRAN mirror from which to install R packages is fixed to a snapshot of CRAN corresponding to the last date for which that version of R was the latest release. These snapshots are provided by the MRAN archive created by Revolution Analytics (now aquired by Microsoft), which archives daily snapshots of all of CRAN, from which a user can install packages with the usual `install.packages()` function  [@MRAN]. Users can always override this default by passing any current CRAN repository explicitly.  Unlike CRAN, Bioconductor only updates its repositories through bi-annual releases aligned to R's Spring release schedule, Bioconductor packages be installed in the usual way using `bioclite`, which automatically selects the Bioconductor release corresponding to the version of R in use.

**Version tags**:  These version tags are propegated throughout this stack: for instance, `rocker/tidyverse:devel` will provide the currently released versions of the R packages in the `tidyverse` [@tidyverse] installed on the nightly build of R-devel.  Developers building packages on this stack are encouraged to tag their images accordingly as well.  Table 3 indicates which versions of R are currently available in the stack, going back to `3.1.0`.  While older versions may be added to the stack at a later date, we note that the MRAN snapshots began in starting 2014-09-17 and thus go back only to R `3.1` era.  Each tag must be built from a separate Dockerfile, enabling minor differences in the build instructions to accomidate changing dependencies.  Dockerfiles for past versions (e.g. < `3.4.0` currently) are intended to remain static over the long term, while the tag for the current version, `latest`, and `devel` may be tweaked to accomidate new features or dependencies.

In this stack, the desired version of R is always built directly from source rather than the `apt` repositories.  Compilers and dependencies are still installed from the stable `apt` repositories, and thus lag behind the more recent versions found in the `testing` stack. Version tags `3.3.3` and older are based on the Debian 8.0 release, codenamed `jessie`, while `3.4.0`, `devel`, and `latest` are based on Debian 9.0, `stretch`, released 2017-06-17, and thus have access to much newer versions of common system dependencies and compilers.  Dependencies needed to compile R that are not required at runtime are removed once R is installed, keeping the base images light-weight for faster download times.  While most system dependencies required by common R packages can still be installed from the `apt` repositories, occassionally a more recent version must be compiled from source (e.g. the Gibbs Sampling library, JAGS, and the geospatial toolkit GDAL, must both be compiled from source on `debian:jessie` images).  In this stack, users should never install R packages, using `apt`, since this will install a second, older version of R from the Debian repositories and a dated version of the R package (since any `r-cran-pkgname` package in the Debian repositories will depend on `r-base` in `apt` as well).

**Build schedule**: All images are built automatically from their corresponding Dockerfiles (found in the GitHub repositories `rocker-org/rocker-versioned` and `rocker-org/geospatial`). A `cron` job sends nightly build triggers to Docker Hub to rebuild the `latest` and `devel` tagged images throughout the stack.  To decrease load on the hub, build triggers for the numeric version tags are sent monthly. Allthough the Dockerfiles for older R versions installs an almost-identical software environment every time, the monthly rebuilding of these images on Docker Hub ensures they continue to recieve Debian security updates from upstream, and proves the build recipe still executes successfully.   Note that rebuilding images with software from external repositories never produces a bitwise identical image, and thus the image identifier hash will change at each build.

image            | description                               | size   | downloads 
---------------- | ----------------------------------------- | ------ | ------- 
[r-ver](https://hub.docker.com/r/rocker/r-ver)          |  Version-stable base R & src build tools  | 219 MB | 2,000
[rstudio](https://hub.docker.com/r/rocker/rstudio)      |  Adds rstudio                             | 334 MB | 285,000
[tidyverse](https://hub.docker.com/r/rocker/tidyverse)  |  Adds tidyverse & devtools                | 656 MB | 14,000
[verse](https://hub.docker.com/r/rocker/verse)          |  Adds java, tex & publishing-related packages   | 947 MB | 4,000
[geospatial](https://hub.docker.com/r/rocker/geospatial)|  Adds geospatial libraries   |  1.3 GB | 1,000

Table: The `rocker-versioned` stack of images

image            | description                               | size   | downloads 
---------------- | ----------------------------------------- | ------ | ------- 
[r-base](https://hub.docker.com/r/_/r-base)        | Version-stable base R & src build tools  | 254 MB | 510,000   
[r-devel](https://hub.docker.com/r/rocker/r-devel) |  R-devel added side-by-side r-base (using alias `RD`)    | 1 GB | 4,000
[r-devel-san](https://hub.docker.com/r/rocker/r-devel-san)  |  as r-devel, but built with compiler sanatizers | 1.1 GB  | 1,000  
[r-devel-ubsan-clan](https://hub.docker.com/r/rocker/r-devel-ubsan-clang)            |  Sanatizers, clang c compiler (instead of gcc) | 1.1 GB | 525
[rstudio:testing](https://hub.docker.com/r/rocker/r-devel-san)        |  rstudio on debian:testing | 1.1 GB | 1,000 
[shiny](https://hub.docker.com/r/rocker/shiny)     |  shiny-server on r-base               | 409 MB | 63,000

Table: the `debian:testing` image stack

tag     | apt repos | MRAN date    | Build frequency  | images with tag 
--------|-----------|--------------|------------------|-------------------------------           
devel   | `stretch` | current date | nigthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
latest  | `stretch` | current date | nightly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.4.0   | `stretch` | current date | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.3.3   | `jessie`  | 2017-04-21   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial` 
3.3.2   | `jessie`  | 2017-03-06   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.3.1   | `jessie`  | 2016-10-31   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
3.3.0   | `jessie`  | 2016-06-21   | monthly          | `r-ver`
3.2.0   | `jessie`  | 2015-06-18   | monthly          | `r-ver`
3.1.0   | `jessie`  | 2014-09-17   | monthly          | `r-ver`

Table: Available tags in the `rocker-versioned` stack.

Customizing builds: `--build-arg` options in `r-ver`: `R_VERSION`, `BUILD_DATE`. in `rstudio`: `RSTUDIO_VERSION`, `PANDOC_TEMPLATES_VERSION`.  
Docker metadata labels: <http://schema-label.org> metadata for `license`, `vcs-url` (GitHub repository), and `vendor`.  These can be altered or extended in downstream images.

## Maintenance and Updates

These images are actively maintained.  This means that while an effort is made to preserve the general function of these images over time, both these Dockerfiles and the resulting images are subject to some change over time.  In particular:

- Images are regularly re-built on Docker Hub whenever their base image changes, starting with changes to `debian` Docker image.  This is the rough equivalent of running `apt-get upgrade` on `debian`, since all `apt-get` commands are re-run and will pull in the most current sources.  This allows the images to receive security updates to any packages installed from the `debian` repositories, but will not in general change the versions of any software and is very unlikely to break anything.

- The Dockerfiles themselves are subject to change, to improve performance, ease of use, readability, or other concerns raised in the issues.  These changes should also not alter the general behavior of R or R packages on the image.  These changes can be seen in the git history.  The [rocker-versioned](https://github.com/rocker-org/rocker-versioned) repo will use its own semantic version tagging to indicate changes to this repository, with snapshots from these tags archived on Zenodo.

## License ##

The Dockerfiles in this repository are licensed under the GPL 2 or later.

# Use cases

@@

@@

This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).
